---
title: "Using R Locally"
description: |
  Setup information for using R on your local computer. 
author:
  - name: Bryan Blanc
date: 01-18-2021
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
---

# R/RStudio Setup

## How to Relocate Local R Package Folder

These steps redefine the R user packages folder in R to use a "Local"
folder within your R user profile, which for me is:
`C:/Users/EshelO/Local/R/win-library/3.4`.

-   The username will need to be modified
-   The R version will need to be modified
-   Any user-writable folder will work; IT has agreed on this as a
    standard location. The same folder can be used for the ArcGIS
    default geodatabase, Github repository directory, etc.).

1.  Relocate your existing R user library folder (which for me was in
    `C:/Users/EshelO/Document/R/win-library/3.4` (if Documents has
    already been migrated, this would be
    `C:/Users/EshelO/OneDrive - Perkins and Will/Documents) to C:/Users/EshelO/Local/R/win-library`
    (or per above)

2.  Test the change:

Run `.libPaths("C:/Users/EshelO/Local/R/win-library/3.4")` (modifying as
necessary) - this is only for your current R session

Run `.libPaths()` and you should be able to validate this worked

Run a test `require()` and `install.packages()` to verify R can both
read and install to the new location

```{r,eval=FALSE}
require(dplyr)  
install.packages("tidycensus") 
```

<!-- 3.  Find your R user directory -->

<!-- -   You can determine your home directory using: `path.expand("~/")` in -->
<!--     R which for me is the Documents folder in my OneDrive\ -->
<!-- -   By default this should be your Documents folder. This would get -->
<!--     relocated to OneDrive, which is fine. -->

<!-- 4.  Update either or both of these files in the folder determined in -->
<!--     (1) - I did both in case some applications only reference one, -->
<!--     although one should suffice: -->

<!-- ```{=html} -->
<!-- <!-- --> -->
<!-- ``` -->
<!-- a)  Run the following line in the R console to change your R library -->
<!--     environment variable -->
<!--     `Sys.setenv(R_LIBS_USER=C:/Users/EshelO/Local/R/win-library/3.4)`\ -->
<!-- b)  add a libpaths setting in the .Rprofile file in your R home -->
<!--     directory: `.libPaths("C:/Users/EshelO/Local/R/win-library/3.4")` -->
<!-- c)  I've attached sample files for both in this folder, if they don't -->
<!--     exist. Modify them to reflect the location where you have your R -->
<!--     user libraries. - You may need to do a Save As to these names, since -->
<!--     you may not be able to rename them to have the leading "." - Use -->
<!--     Notepad++ if you have it. Notepad also works; you may need to set -->
<!--     file type as All Files so it does not append ".txt" -->
<!-- d)  After restarting run .libPaths() and/or  Sys.getenv() and you should -->
<!--     be able to validate this worked (for the latter check the -->
<!--     R_LIBS_USER setting) and do the same require and packages.install() -->
<!--     tests as above. -->

# Other  Setup

## Installing and Configuring Git & GitHub

Git is a version control system, a tool that tracks changes to your code
and shares those changes with others. Git is most useful when combined
with GitHub, a website that allows you to share your code with the
world, solicit improvements via pull requests and track issues. Git +
GitHub is the most popular version control system for developers of R
packages (thousands of R packages are hosted on GitHub) and many other
types of software.

Git and GitHub are generally useful for all software development and
data analysis, not just R packages.

To use Git/GitHub with R, you will need to set up a GitHub account and
configure RStudio to interface with GitHub. Detailed instructions for
this can be found [here](http://r-pkgs.had.co.nz/git.html). 

File a help ticket with IT to have Git installed. 

**- To get an Enterprise P+W account:** - Create your own personal
github account; recommend setting up using your personal e-mail - Email
[Joel Register](mailto:joel.register@perkinswill.com) your github
username and ask to be added to the Nelson\Nygaard team - "Could I
request to join the enterprise GitHub account? My username is XXXXX."

## Installing RTools

RTools is a collection of development tools that are essential for
writing and releasing packages. The latest RTools package can be
downloaded from the [CRAN
website](https://cran.r-project.org/bin/windows/Rtools/). File a help ticket to have it installed. If everything goes well, you
will have a new directory on your computer called C:/Rtools. More
detailed information and instructions can be found
[here](https://cran.r-project.org/doc/manuals/R-admin.html#The-Windows-toolset).

# Script Management

## Folder Conventions

-   Create R folder within G: Drive

    -   Talk to GIS team about adding this to the template (Oren to
        discuss)

-   Folder structure

    -   README.txt

        -   Title block for code with author, date, purpose

    -   Add shortcuts to P - Analysis

    -   Esther suggests three subfolders:

        -   Raw: shortcuts to G - z.orig
        -   Interim
        -   Output: shortcut to P -- Graphics or SharePoint -- Graphics
            folders

-   Start using R Projects?

    -   Allows relative file paths

## Script Titleblock Snippet

In R studio go to Tools -\> global options --\> code --\> bottom of page
"edit snippets" --\> scroll to bottom and paste the below in, change
your name in the Author field.

```{r, eval=FALSE}
snippet header
	## ---------------------------
	##
	## Script name: 
	##
	## Purpose of script:
	##
	## Project Name: 
	##
	## Project Number: 
	##
	## Author: Esther Needham
	##
	## Date Created: `r paste(Sys.Date())`
	##
	## ---------------------------
	##
	## Notes:
	##   
	##
	## ---------------------------
	## set working directory here or make sure there is a .Rproj file
	## ---------------------------
	options(scipen = 6, digits = 4) # view outputs in non-scientific notation
	## ---------------------------
	## load up the packages we will need:  (uncomment as required)
	require(tidyverse)
	require(sf)
	## ---------------------------
	## load up functions from other scripts into memory
	# source("functions/summarise_data.R") #example
	## ---------------------------
```

# ShinyApps Web Hosting

## Management Dashboard

-   <https://www.shinyapps.io/admin/#/profile>

## Management Account

-   [shinyapps\@nelsonnygaard.com](mailto:shinyapps@nelsonnygaard.com){.email}

-   This account is accessible to various people (Bryan, Tomoko, Esther,
    Oren)

-   You can add it to your email by:

    -   Outlook -\> Accounting Settings
    -   Email Tab -\> New
    -   Enter
        [shinyapps\@nelsonnygaard.com](mailto:shinyapps@nelsonnygaard.com){.email}
        -\> Next
    -   Restart Outlook and you should see the mailbox at the bottom of
        the left navigation toolbar.

-   Email is also forwarded to each of the above individuals. Forwarding
    is set up using a server side rule by logging into Outlook 365 web
    mail. If others need to be added, this can be done by:

    -   Log into [Outlook
        365](https://outlook.office.com/mail/shinyapps@nelsonnygaard.com/inbox)
        (this link should take you straight to the account)

        -   If not, go to icon in upper right -\> Open Another Mailbox
            -\>
            [Shinyapps\@nelsonygaard.com](mailto:Shinyapps@nelsonygaard.com){.email}

    -   Settings icon in upper right of Outlook

    -   Mail settings -\> Rules

    -   "Shinyapps Email Forwarding" Rule -\> Edit Rule

### Password

-   Oren requested that this account be added to Okta, which hopefully
    can then be used to save our password.

### Uploading to shinyapps.io

-   Use rsconnect package in RStudio to deploy using a token in the
    Shiny dashboard
-   In Shiny Dashboard, go to Account -\> tokens -\> Show -\> Copy to
    Clipboard
-   One-time: paste the commend from the Shiny Dashboard --
    rsconnect::setaccount...\
-   In RStudio, run: rsconnect::deployApp()
-   Set working directory to where Shiny app lives
-   Setwd... (may need to comment and uncomment)

## AWS SQL Server Hosting

-   How do we charge costs do a credit card? Bryan needs to hook in a
    credit card for this.

## Google API Access

-   Individual accounts needed, threshold beyond (\~2500 per day) which
    you are billed per query
-   Action: re-investigate central account/billing -- Oren and/or Tomoko
