---
title: "Nelson\\\\Nygaard's R Training Home Page"
author: "Bryan Blanc"
date: "Updated as of `r strftime(Sys.Date(),'%B %d, %Y')`"
output: 
  html_document:
    toc: true
---

```{r, include = FALSE, warning=FALSE, message=FALSE}
require(tidyverse)
require(dplyr)
require(readxl)
require(lubridate)
require(janitor)
require(hms)
require(DT)
```

## Purpose

Bryan Blanc has been re-hired on a temporary basis (TBD long-term) for a three month appointment, through October 13th, 2020. One of the main goals of this appointment is to train other Nelson\\Nygaard employees in using R, particularly the use of Shiny for web-based dashboards. 

This document forms the work plan and content repository for Bryan's R training. The training will be structured as a 10 week 'course', which will include a ~1 hour webinar each week for 10 weeks beginning on the week of August 10th. The below topics are proposed for the course, and can be edited based on feedback. The content for each webinar (including narrative and sample code) will live in this website (in HTML files), or in the case of the Shiny apps themselves, on the Nelson\\Nygaard shinyapps.io server (with code stored on GitHub). All content will be linked in this document.

Shiny will be a recurring point of discussion during the course, with four of the ten webinars devoted exclusively to Shiny. People attending the Shiny webinars (identified in the schedule below) will be expected to identify a project of theirs with data that lends itself to a Shiny dashboard by the time of the first Shiny webinar (week of August 9th). Participants will work with their team leaders the relevant project manager to identify a project and decide if some of the time they spend on developing an example Shiny app is billable. 

## Schedule

```{r,echo=FALSE, warning=FALSE, message=FALSE,out.width=1500}
schedule = read_excel('../data/nn_r_info_db.xlsx') %>%
  mutate(webinar_time = strftime(as_hms(webinar_time),'%I:%M %p'),
         webinar_date = strftime(force_tz(webinar_date,tz='America/Los_Angeles'),'%B %d, %Y'),
         week_start_date = strftime(force_tz(week_start_date,tz='America/Los_Angeles'),'%B %d, %Y'),
         week_end_date = strftime(force_tz(week_end_date,tz='America/Los_Angeles'),'%B %d, %Y'),
         webinar_topic = ifelse(!is.na(topic_link),
                                paste0("<a href='",topic_link,"'>",webinar_topic,"</a>"),
                                webinar_topic),
         recording_link = ifelse(!is.na(recording_link),
                                paste0("<a href='",recording_link,"'>",'Recording',"</a>"),
                                recording_link)) %>%
  dplyr::select(-topic_link,-articles,-books,-datacamp_courses,-hrs_available,
                -week_end_date) %>%
  rename(`Week #`=week_num,
         `Category`=category,
         `Week Start Date`=week_start_date,
         `Webinar Topic`=webinar_topic,
         `Webinar Date`=webinar_date,
         `Webinar Time`=webinar_time,
         `Recording Link`=recording_link)

datatable(schedule,
          rownames=FALSE, escape=FALSE,
          options = list(searching=FALSE,paging=FALSE,info=FALSE,
                         autoWidth=TRUE,
                         columnDefs = list(list(width = '150px', targets = c(2)))),
          width='100%')

```

## Topics

### [Tidyverse](topics_output/tidyverse.html)

The [Tidyverse](https://www.tidyverse.org/) is a 'universe' of packages designed for data science with R. They are all built with an underlying design philosophy and grammar based on the [concept of 'tidy' data](https://vita.had.co.nz/papers/tidy-data.pdf). This session will discuss the tidy data concept and its main package implementations. For me, the tidy data concept influences not just the work I do in R but how I structure spreadsheets and datasets in general, and I have found it to generally result in cleaner data analyses.[Tidyverse packages](https://www.tidyverse.org/packages/) are among the most used R packages, and greatly improve upon implementations of basic operations in base R. The Tidyverse chapter will greatly draw upon Hadley Wickham's (the progenitor of the Tidyverse) book (available for free online), [R for Data Science](https://r4ds.had.co.nz/).

### Shiny

#### [Shiny Session 1](topics_output/shiny_1.html)

This session will discuss the basics of Shiny:

- What is Shiny? 
- How is a Shiny app structured?
- How do I write a basic Shiny application?

Please come prepared with an idea of what project you will develop a basic Shiny app for over the next eight weeks. Work with your team leader and the relevant project manager to identify a project and decide if some of the time you spend on developing an example Shiny app is billable.  You shouldn't have to spend more than a few hours in advance of each of the four webinars -- these apps can be very basic. Progress is expected for discussion from each participant each session. 

#### [Shiny Session 2](topics_output/shiny_2.html)

This session will discuss some more features of Shiny, namely:
- Loading data and functions
- Using `leaflet` with Shiny
- Using the specialized package for dashboards, `shinydashboard`. 

We will also discuss everyone's proposed demo application. 

#### [Shiny Session 3](topics_output/shiny_3.html)

#### [Shiny Session 4](topics_output/shiny_4.html)

### [Geospatial Data](topics_output/geospatial.html)

This module discusses how to use the `sf` package for geospatial data and how to construct interactive maps using `leaflet`. 

### [Census, TIGER, and LODES data](topics_output/census.html)

### [GTFS data](topics_output/gtfs_data.html)

### [Misc. Questions](topics_output/misc_questions.html)

This session will provide an opportunity to ask any questions of Bryan related to R that were not covered in prior modules. Please submit questions through this [Google Form](https://docs.google.com/forms/d/e/1FAIpQLSc6JMTAwDHN82RASgRUYJggEeRri67WH1euaoxNFtjgNxboqg/viewform?usp=sf_link) throughout the course that don't appear to be answered in any of the other course modules, Bryan will curate the questions into a ~1 hour presentation for this final module. Stay tuned for the presentation itself, it will be prepared before the final week. 

## Other Topics Not Included in Course

### [R Setup at Nelson\\Nygaard](other_topics/r_setup_nn.html)

Esther Needham, Oren Eshel, and Paul Leitman compiled some standards for R setup at Nelson\\Nygaard. That content is covered in [this article](other_topics/r_setup_nn.html). 

### Git and GitHub

Esther Needham did a presentation to Nelson\\Nygaard staff on using Git and GitHub for R project work. A PDF of that presentation is available [here](docs/PDF/NN GitHub 101_ Git-ting Started with Git and GitHub.pdf). 

### R Markdown

R Markdown is a great way to produce a cohesive presentation of results from your R code -- in case you were wondering how this website was produced, it was produced with 🤯R Markdown🤯! R Markdown is a R implementation of a popular markup language -- Markdown -- that allows for the seamless intermingling of text, R code, and results (e.g., plots, tables, HTML widgets like leaflet maps) that can be rendered into formats like PDF and HTML (other formats include Microsoft Word -- for a full list see [here](https://rmarkdown.rstudio.com/lesson-9.html))

