---
title: "R Setup at Nelson\\\\Nygaard"
author: "Esther Needham, Oren Eshel, Paul Leitman, and Bryan Blanc"
date: "7/29/2020"
output: 
  html_document:
    toc: true
    toc_depth: 4
---

[Back to overview](../index.html)

# Neccessary Setup 

## Summary 

The table below summarizes the software packages needed to run R. Installation files for use by IT are saved here: O:/software tools/R/Resources/Installation Files

| Software Package   | Recommended Version   | Download Link                                      |
|--------------------|-----------------------|----------------------------------------------------|
| R                  | 4.0.1                 | https://www.r-project.org/                         |
| RStudio            | 1.3.959               | https://www.rstudio.com/products/rstudio/download/ |
| RTools             | 40                    | https://cran.r-project.org/bin/windows/Rtools/     |
| Git                | 2.27.0                | https://git-scm.com/download/win                   |

## Installing R

You’ll first need to install R itself (the software environment) in order to use it.  

**Install Link:** https://cran.r-project.org/bin/windows/base/ 

## Installing RStudio 

The basic R installation only comes with a very minimal console, so you’ll also want to install RStudio. RStudio is an Integrated Development Environment (IDE) produced and maintained by RStudio (a for-profit company).  

**Install Link:** https://www.rstudio.com/products/rstudio/download/ 

## How to Relocate Local R Package Folder

These steps redefine the R user packages folder in R to use a "Local" folder within your R user profile, which for me is: "C:/Users/EshelO/Local/R/win-library/3.4". 

- The username will need to be modified 
- The R version will need to be modified 
- Any user-writable folder will work; IT has agreed on this as a standard location. The same folder can be used for the ArcGIS default geodatabase, Github repository directory, etc.). 

1. Relocate your existing R user library folder (which for me was in C:/Users/EshelO/Document/R/win-library/3.4 (if Documents has already been migrated, this would be C:/Users/EshelO/OneDrive - Perkins and Will/Documents) to C:/Users/EshelO/Local/R/win-library (or per above)  

2. Test the change:  

Run `.libPaths("C:/Users/EshelO/Local/R/win-library/3.4")` (modifying as necessary) - this is only for your current R session  

Run `.libPaths()` and you should be able to validate this worked  

Run a test `require()` and `install.packages()` to verify R can both read and install to the new location  

```{r,eval=FALSE}
require(dplyr)  
install.packages("tidycensus”) 
```

3. Find your R user directory 
  - You can determine your home directory using: path.expand("~/") in R which for me is the Documents folder in my OneDrive  
  - By default this should be your Documents folder. This would get relocated to OneDrive, which is fine. 

4. Update either or both of these files in the folder determined in (1) - I did both in case some applications only reference one, although one should suffice:

  a) Run the following line in the R console to change your R library environment variable `Sys.setenv(R_LIBS_USER=C:/Users/EshelO/Local/R/win-library/3.4)`  
  b) add a libpaths setting in the .Rprofile file in your R home directory: `.libPaths("C:/Users/EshelO/Local/R/win-library/3.4")`
  c) I've attached sample files for both in this folder, if they don't exist. Modify them to reflect the location where you have your R user libraries. 
    - You may need to do a Save As to these names, since you may not be able to rename them to have the leading “.” 
    - Use Notepad++ if you have it. Notepad also works; you may need to set file type as All Files so it does not append “.txt” 
  d) After restarting run .libPaths() and/or  Sys.getenv() and you should be able to validate this worked (for the latter check the R_LIBS_USER setting) and do the same require and packages.install() tests as above. 

# Optional Setup

## Installing and Configuring Git & GitHub
Git is a version control system, a tool that tracks changes to your code and shares those changes with others. Git is most useful when combined with GitHub, a website that allows you to share your code with the world, solicit improvements via pull requests and track issues. Git + GitHub is the most popular version control system for developers of R packages (thousands of R packages are hosted on GitHub) and many other types of software. 

Git and GitHub are generally useful for all software development and data analysis, not just R packages. 

To use Git/GitHub with R, you will need to set up a GitHub account and configure RStudio to interface with GitHub. Detailed instructions for this can be found [here](http://r-pkgs.had.co.nz/git.html). 

**- To get an Enterprise P+W account:**
  - Create your own personal github account; recommend setting up using your personal e-mail
  - Email [Joel Register](mailto:joel.register@perkinswill.com) your github username and ask to be added to the Nelson\Nygaard team 
    - “Could I request to join the enterprise GitHub account? My username is XXXXX.”  
  - Uses Git; Esther suggests using command-line 
  - Integration with RStudio also an option 
  - Follow up brownbag (Esther action?) 
  - 3/25/2020: Esther has developed resources; see PPT file and Wiki page 

## Installing RTools 

RTools is a collection of development tools that are essential for writing and releasing packages.  The latest RTools  package can be  downloaded from the [CRAN website](https://cran.r-project.org/bin/windows/Rtools/).  Download the latest package and run the executable file. If everything goes well, you will have a new directory on your computer called C:/Rtools. More detailed information and instructions can be found [here](https://cran.r-project.org/doc/manuals/R-admin.html#The-Windows-toolset). 

# Script Management

## Folder Conventions

- Create R folder within G: Drive 
  - Talk to GIS team about adding this to the template (Oren to discuss) 
- Folder structure 
  - README.txt 
    - Title block for code with author, date, purpose 
  - Add shortcuts to P - Analysis 
  - Esther suggests three subfolders: 
    - Raw: shortcuts to G - z.orig 
    - Interim 
    - Output: shortcut to P – Graphics or SharePoint – Graphics folders 
- Start using R Projects? 
  - Allows relative file paths 
  
## Script Titleblock Snippet

In R studio go to Tools -> global options --> code --> bottom of page "edit snippets" --> scroll to bottom and paste the below in, change your name in the Author field.

```{r, eval=FALSE}
snippet header
	## ---------------------------
	##
	## Script name: 
	##
	## Purpose of script:
	##
	## Project Name: 
	##
	## Project Number: 
	##
	## Author: Esther Needham
	##
	## Date Created: `r paste(Sys.Date())`
	##
	## ---------------------------
	##
	## Notes:
	##   
	##
	## ---------------------------
	## set working directory here or make sure there is a .Rproj file
	## ---------------------------
	options(scipen = 6, digits = 4) # view outputs in non-scientific notation
	## ---------------------------
	## load up the packages we will need:  (uncomment as required)
	require(tidyverse)
	require(sf)
	## ---------------------------
	## load up functions from other scripts into memory
	# source("functions/summarise_data.R") #example
	## ---------------------------
```
  
# ShinyApps Web Hosting

## Management Dashboard

- [https://www.shinyapps.io/admin/#/profile](https://www.shinyapps.io/admin/#/profile)

## Management Account

- shinyapps@nelsonnygaard.com
- This account is accessible to various people (Bryan, Tomoko, Esther, Oren)
- You can add it to your email by: 
  - Outlook -> Accounting Settings 
  - Email Tab -> New 
  - Enter shinyapps@nelsonnygaard.com -> Next 
  - Restart Outlook and you should see the mailbox at the bottom of the left navigation toolbar. 
- Email is also forwarded to each of the above individuals. Forwarding is set up using a server side rule by logging into Outlook 365 web mail. If others need to be added, this can be done by: 
  - Log into [Outlook 365](https://outlook.office.com/mail/shinyapps@nelsonnygaard.com/inbox) (this link should take you straight to the account) 
    - If not, go to icon in upper right -> Open Another Mailbox -> Shinyapps@nelsonygaard.com 
  - Settings icon in upper right of Outlook 
  - Mail settings -> Rules 
  - “Shinyapps Email Forwarding” Rule -> Edit Rule 
  
### Password

- Oren requested that this account be added to Okta, which hopefully can then be used to save our password. 

### Uploading to shinyapps.io

- Use rsconnect package in RStudio to deploy using a token in the Shiny dashboard 
- In Shiny Dashboard, go to Account -> tokens -> Show -> Copy to Clipboard 
- One-time: paste the commend from the Shiny Dashboard – rsconnect::setaccount…  
- In RStudio, run: rsconnect::deployApp() 
- Set working directory to where Shiny app lives 
- Setwd… (may need to comment and uncomment) 

## AWS SQL Server Hosting

- How do we charge costs do a credit card? Bryan needs to hook in a credit card for this.  

## Google API Access

- Individual accounts needed, threshold beyond (~2500 per day) which you are billed per query 
- Action: re-investigate central account/billing – Oren and/or Tomoko 